# The Building Blocks of `{ggplot2}`

In this chapter, you will begin to learn about using the `{ggplot2}` functions to visualize your data. While there are multiple ways to plot data in R, the functions from ggplot2 are not only simpler to use than base R, but it also produces far more aesthetically pleasing plots.

:::fyi
The ggplot2 functions are all available when we load the `{tidyverse}` package.
:::

<br />


## ggplot2

The `gg` in `{ggplot2`} stands for *grammar of graphics*. The grammar of graphics [@wilkinson:2005] is a formal system of expressive grammatical rules for creating perceivable graphs. The grammar of graphics includes things such as aesthetics, geometries, scales, facets, and guides. Hadley Wickham adopted this grammar into the initial `{ggplot}` package, which he then re-wrote and updated to create the `{ggplot2}` package. As you learn how to create plots using `{ggplot2}`, you will also begin to learn the grammar of graphics. Understanding this grammar will help you describe, and create almost any visualization you can imagine.

:::fyi
**IMPORTANT**

As you read the remainder of this chapter do not worry about the specific syntax/code. Instead focus on the ideas that are being presented. You will learn the specific code later as we encounter it in class.
:::


<br />


## Understanding the Basic Syntax

We will use the **gapminder.csv** data to illustrate how to create and modify visualizations using functionality in `{ggplot2}`. As a reminder, in the first code chunk of your script file, load the `{tidyverse}` library. In a second code chunk, use the `read_csv()` function to import the *gapminder.csv* data and assign the data into an object called `gapminder`. (You also might want to view the data to ensure that it imported into RStudio correctly.)



```{r}
#| label: load-packages2
#| message: false

######################
# Load libraries
######################

library(tidyverse)


######################
# Import data
######################

gapminder <- read_csv("data/gapminder.csv")

# View data
gapminder
```



Plots in `{ggplot2`} are built by *layering* different components. For example, consider the following syntax which creates a scatterplot using the income and CO2 emissions attributes:

```{r}
#| label: fig-understand
#| fig-cap: "Scatterplot of CO2 emissions versus income."
#| fig-alt: "Scatterplot of CO2 emissions versus income."
ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_point() +
  geom_smooth()
```

In this syntax there are three layers used to create the plot:

1.  a global layer,
2.  a geometric layer drawing points, and
3.  a geometric layer drawing a smoother.

The layers are literally *summed* together with the `+` operator to form the plot. We will look at each of these layers in turn.

<br />


### The Global Layer

The first layer (referred to as the global layer) in every plot you create employs the function `ggplot()`. It contains a reference to the source data (`data=`) and any *global aesthetic mappings* (more about this later). The first layer only sets up the plot, it doesn't actually plot anything.

```{r}
#| label: fig-global
#| fig-cap: "The background layer for the plot is drawn. The domain (x-values) and range (y-values) is based on data in the income and co2 attributes."
#| fig-alt: "The background layer for the plot is drawn. The domain (x-values) and range (y-values) is based on data in the income and co2 attributes."

##########################
# Running the global layer
##########################

ggplot(data = gapminder, aes(x = income, y = co2))
```

-   The `data=` argument indicates the source data frame.
-   The `aes=` argument sets the aesthetic mapping(s).

<br />


### Aesthetics and Aesthetic Mappings

Aesthetics define what we, as humans, perceive in a given plot; that is the visual properties of a plot. For example, the position of certain elements in the plot (where they are located), or the color or transparency of an element. These aesthetics can be fixed or variable. For example, consider the color of a set of points. If the color is fixed, it would be the same for all the points. If the color is variable, it might be red for some points and blue for other points. Often, this variability in the aesthetic is based on some attribute in our dataset (e.g., points representing Democrats are colored blue, while those representing Republicans are colored red).

Linking an aesthetic to an attribute in the data, is referred to as an *aesthetic mapping*. That is because aesthetic mappings map visual properties in the plot (e.g., position, color) to the values in a particular attributes in the data. Aesthetic mappings are specified in an `aes()` function. In our earlier example syntax, there were two aesthetic mappings that were defined:

1.  Income values from the data will be mapped to the *x*-position.
2.  CO2 values from the data will be mapped to the *y*-position.

These mappings were used to define the domain (*x*-values) and range (*y*-values) for the blank plot created by the `ggplot()` global layer.

:::protip
**IMPORTANT**

All aesthetic mappings need to be specified in an `aes()` function. Because aesthetic mappings use information in the attribute to apply the specified visual properties, the aesthetics (e.g., `x=`, `y=`) need to be set to an attribute name in the data frame. For example, both `income` and `co2` are attribute names in the `gapminder` data object.
:::

<br />


### Adding Geometric Layers

After starting with the global layer, we can modify our visualization by adding layers to the global layer. For example, the layer that includes the function `geom_point()` is being added to the global layer in the syntax below.

```{r}
#| label: fig-adding
#| fig-cap: "Points are added on top of the plot's background layer."
#| fig-alt: "Points are added on top of the plot's background layer."

##########################
# Global layer + geometric layer adding points
##########################

ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_point()
```

Geometric layers (`geom_*`) are used to add things like points, lines, histograms, densities, etc. The `geom_point()` layer in our plot is actually drawing the points of the scatter plot. These layers draw on the aesthetic mappings defined in the global layer to know where to draw these geometric objects. For example, the data in the `income` and `co2` variables define the *x*- and *y*-positions of the points, respectively.

When layers are added they are "stacked" on top of previous layers. For example, consider the two sets of syntax below. Each will again create our scatterplot by adding a set of points based on the data in the global layer. Each will also draw a  smoother (a sort of trend line). However, in the first set of syntax, the smoother is drawn on top of the points, and in the second set of syntax, the points are drawn on top of the smoother.

```{r}
#| eval: false

##########################
# LEFT: Draw smoother on top of points
##########################
ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_point() +
  geom_smooth()
```

```{r}
#| eval: false

##########################
# RIGHT: Draw points on top of smoother
##########################

ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_smooth() +
  geom_point()
```

```{r}
#| label: fig-layer
#| fig-cap: "LEFT: The smoother is drawn on top of the points. RIGHT: The points are drawn on top of the smoother."
#| fig-alt: "LEFT: The smoother is drawn on top of the points. RIGHT: The points are drawn on top of the smoother."
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| out-width: "100%"

# Load library
library(patchwork)

# Draw smoother on top of points
p1 = ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_point() +
  geom_smooth(alpha = 1)

# Draw points on top of smoother
p2 = ggplot(data = gapminder, aes(x = income, y = co2)) + 
  geom_smooth(alpha = 1) +
  geom_point()

p1 | p2
```

A useful metaphor might be to imagine putting together a collage of many different photographs. Some parts of some of the photographs (especially those you put down first) might be covered up by photographs that you add later.

:::protip
**PROTIP**

When you are creating plots, you might need to switch the order of some of the layers to get the visualization you want.
:::

<br />


<!-- ### Adding the Theme Layer -->

<!-- The third layer in our initial scatterplot (`theme_minimal()`) is referred to as a theme layer. It changes the entire theme of the plot (e.g., color scheme, font, axes, gridlines).  -->

<!-- ```{r} -->
<!-- #| label: fig-layer -->
<!-- #| fig-cap: "LEFT: The default grey theme. RIGHT: The minimal theme." -->
<!-- #| fig-alt: "LEFT: The default grey theme. RIGHT: The minimal theme." -->
<!-- #| fig-width: 6 -->
<!-- #| fig-height: 6 -->
<!-- #| fig-show: "hold" -->
<!-- #| out-width: "45%" -->

<!-- ########################## -->
<!-- # Plot with the default theme -->
<!-- ########################## -->

<!-- ggplot(data = gapminder, aes(x = income, y = co2)) +  -->
<!--   geom_point() -->

<!-- ########################## -->
<!-- # Plot with the minimal theme -->
<!-- ########################## -->

<!-- ggplot(data = gapminder, aes(x = income, y = co2)) +  -->
<!--   geom_point() + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- <br /> -->


## Good Syntactic Habits

As we add multiple layers to build up our plot, it is a good habit to use multiple lines for the syntax. Generally we put one layer on each line.

```{r eval=FALSE}
##########################
# Multiple layers -- One on each line
##########################

ggplot(data = gapminder, aes(x = income, y = co2)) +  
  geom_smooth() +
  geom_point()
```

The `+` sign needs to be at the end of the line (not at the beginning). If you are using a script file (which you should be), highlight ALL layers in the plot and click `Run` to create the plot.

:::protip
**PROTIP**

If you write your `{ggplot2}` syntax across multiple lines, it makes it not only easier to read, but also easier to try out different layers or features. For example, you can comment out lines to remove layers as you are building the plot without losing the syntax you wrote.

You can also debug syntax by highlighting all lines up to a given `+` sign and running the syntax. By subsequently highlighting and running additional lines of syntax, you can often figure out where any errors show up!
:::

<br />


<!-- ## Global vs. Local Aesthetic Mappings -->

<!-- As you learned earlier, we use the `aes()` function to set aesthetic mappings (i.e., linking attributes to aesthetics). Aesthetic mappings are often set globally (in the initial `ggplot()` layer). When we include aesthetic mappings in the global layer they are applied to *all layers* in the plot. For example, the syntax below maps data in the `income` attribute to the *x*-position, data from the `co2` attribute to the *y*-position, and data from the `region` attribute to color for both the loess smoother and the points. -->

<!-- ```{r} -->
<!-- #| label: fig-aes -->
<!-- #| fig-cap: "The global aesthetic mappings of x- and y-positioning, as well as, color are applied to the smoother and point layers." -->
<!-- #| fig-alt: "The global aesthetic mappings of x- and y-positioning, as well as, color are applied to the smoother and point layers." -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 6 -->
<!-- #| out-width: '80%' -->

<!-- ########################## -->
<!-- # All aesthetic mappings are global -->
<!-- ########################## -->

<!-- ggplot(data = gapminder, aes(x = income, y = co2, color = region)) +   -->
<!--   geom_smooth() + -->
<!--   geom_point() -->
<!-- ``` -->

<!-- :::protip -->
<!-- **FYI** -->

<!-- Notice when we use non-positional aesthetic mappings (e.g., color) a legend, or guide in the grammar of graphics, will be automatically added to our plot for each aesthetic mapping. -->
<!-- ::: -->

<!-- Aesthetic mappings can also be set locally in a specific layer. Aesthetic mappings set in a specific layer only apply to that particular layer. Below, we continue to globally map data from the `income` attribute to the *x*-position, and data from the `co2` attribute to the *y*-position. Both the smoother layer and point layer will utilize that. However, only the point layer will map data from the `region` attribute to color. Note that regardless of whether the aesthetic mapping is global or local, we still make this mapping inside an `aes()` function. -->

<!-- ```{r} -->
<!-- #| label: fig-aes2 -->
<!-- #| fig-cap: "The global aesthetic mappings of x- and y-positioning are applied to the smoother and point layers. The aesthetic mapping of color is only applied to the point layer." -->
<!-- #| fig-alt: "The global aesthetic mappings of x- and y-positioning are applied to the smoother and point layers. The aesthetic mapping of color is only applied to the point layer." -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 6 -->
<!-- #| out-width: '80%' -->

<!-- ########################## -->
<!-- # Aesthetic mappings for x- and y-position are global -->
<!-- # Aesthetic mapping for color only applies to the points layer -->
<!-- ########################## -->

<!-- ggplot(data = gapminder, aes(x = income, y = co2)) +   -->
<!--   geom_smooth() + -->
<!--   geom_point(aes(color = region)) -->
<!-- ``` -->

<!-- <br /> -->


<!-- ## Fixed Aesthetics -->

<!-- Fixed aesthetics assign the exact same value for the visual property for **all** the observations; it is not based on the data. Fixed aesthetics do not go inside of the `aes()` function. (Remember: Inside of the `aes()` function the mapping has to be to an attribute name!) The following syntax uses the following aesthetics: -->

<!-- -   **Global aesthetic mapping:** Data from the `income` attribute mapped to the *x*-position. -->
<!-- -   **Global aesthetic mapping:** Data from the `co2` attribute mapped to the *y*-position. -->
<!-- -   **Fixed local aesthetic:** All points are given a shape of `22` (filled square with a border). You can see all the shape options [here](http://www.sthda.com/english/wiki/ggplot2-point-shapes). -->
<!-- -   **Fixed local aesthetic:** All points are given a size of `4` (slightly bigger). The default size is 3. -->
<!-- -   **Local aesthetic mapping:** Data from the `region` attribute mapped to fill for the points layer only. Because we are using a different shape we use `fill=` rather than `color=` to color the observations. -->
<!-- -   **Fixed local aesthetic:** All points are given a border color of black (for this shape border color is set using `color=`). -->

<!-- ```{r} -->
<!-- #| label: fig-aes3 -->
<!-- #| fig-cap: "The global aesthetic mappings of x- and y-positioning are applied to the smoother and point layers. The aesthetic mapping of color is only applied to the point layer. The point layer also includes several fixed aesthetics." -->
<!-- #| fig-alt: "The global aesthetic mappings of x- and y-positioning are applied to the smoother and point layers. The aesthetic mapping of color is only applied to the point layer. The point layer also includes several fixed aesthetics." -->
<!-- #| fig-width: 8 -->
<!-- #| fig-height: 6 -->
<!-- #| out-width: '80%' -->

<!-- ########################## -->
<!-- # Aesthetic mappings for x- and y-position are global -->
<!-- # Aesthetic mapping for color only applies to the points layer -->
<!-- # All points are the same shape, size, and border color -->
<!-- ########################## -->

<!-- ggplot(data = gapminder, aes(x = income, y = co2)) +   -->
<!--   geom_smooth() + -->
<!--   geom_point(aes(fill = region), shape = 22, size = 4, color = "black") -->
<!-- ``` -->

<!-- <br /> -->



## References