# Creating Scatterplots in R


This chapter will focus on how we create scatterplots in R. To illustrate this, we will use the **world-happiness.csv** data.  This data includes 2019 indices of happiness, wealth, and life expectancy for 156 countries.

To create a scatterplot, we will use functions from the `{tidyverse}` library. As a reminder, in the first code chunk of your script file, load the `{tidyverse}` library. In a second code chunk use the `read_csv()` function to import the *world-happiness.csv* data and assign the data into an object called `happy`. You also might want to view the data in the data viewer to ensure that it imported into RStudio correctly. (Don't forget to add code comments as well!)


```{r}
#| label: load-packages2
#| message: false

######################
# Import data
######################

library(tidyverse)


######################
# Import data
######################

happy <- read_csv("data/world-happiness.csv")
```


To create a scatterplot to visualize the relationship between life expectancy and happiness, we will first create a new code chunk in your script file. In the ggplot syntax we will use two layers in our plot: `ggplot()` and `geom_point()`. Remember that the `ggplot()` layer (the global layer) indicates the data object and sets the primary aesthetic mapping for the plot. The `geom_point()` layer will use the information in the global layer to create the scatterplot.

In the global layer we will indicate the data object is `happy` using the argument `data=happy`. In a scatterplot, we also set the aesthetic mapping for the *x*- and *y*-positions using `x=` and `y=` in the `aes()` function. Here we will map life expectancy to the *x*-coordinate and happiness to the *y*-coordinate. The syntax is as follows:

```{r}
#| label: fig-scat-01
#| fig-cap: "Scatterplot displaying the relationship between life expectancy and happiness."
#| fig-alt: "Scatterplot displaying the relationship between life expectancy and happiness."
#| out-width: "90%"
#| message: false

#######################################
# Create scatterplot
#######################################

ggplot(data = happy, aes(x = life_exp, y = happiness)) +
  geom_point()
```

<br />


## Changing the Point Shape, Size, Color and Fill

The default point shape for the scatterplot is a small, filled in circle. This shape has a value of 19 (see below). So in the `geom_point()` layer it is implicitly using `geom_point(shape=19)`. Here are all of the point shapes and their corresponding values:

```{r}
#| label: fig-shapes
#| fig-cap: "Point shapes and their corresponding values."
#| fig-alt: ""
#| echo: false

pch_ref <- data.frame(
  shape = 0:25,
  x = rep(0:6, length.out = 26),
  y = -rep(0:3, each = 7, length.out = 26)
)
ggplot(pch_ref, aes(x = x, y = y)) +
  geom_point(aes(shape = shape), size = 5, colour = "#3a86d4",
             fill = "grey80") +
  geom_text(aes(label = shape), vjust = 2.4, size = 3.5) +
  scale_shape_identity() +
  labs(title = "ggplot2 point shapes (shape= 0–25)") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        panel.grid = element_blank())
```

Each of these shapes has different possibilities for `color=` and `fill=`. For example, with the default shape of 19, there is no border on the point and to change the point color you would use `color=`. In contrast, for shape 21 (which is also a circle), `color=` would change the border color and `fill=` would chnage the interior color. 

Here we change the point shape to 21 and add both a border and interior color for the points. We also make the points larger by adding `size=` to the `geom_point()` layer. (The default size is 1.5.) We also add labels to our plot. 


```{r}
#| fig-alt: "Scatterplot displaying the relationship between life expectancy and happiness."
#| out-width: "90%"
#| message: false

#######################################
# Create scatterplot
# Change point shape, color, fill, and size
#######################################

ggplot(data = happy, aes(x = life_exp, y = happiness)) +
  geom_point(shape = 21, color = "black", fill = "#ffde7a", size = 3) +
  labs(
    title = "LIFE EXPECTANCY AND HAPPINESS",
    subtitle = "\nCountries with higher life expectancies tend to be happier.",
    caption = "SOURCE: 2019 World Happiness Report",
    x = "Life expectancy (in years)",
    y = "Happiness (on a scale from 0-10)"
  )
```


<br />


## Axis Limits

To change the axis limits, we will use the `scale_x_continuous()` and `scale_y_continuous()` layers. (Remember we use the continuous scales since the attributes mapped to *x* and *y* are quantitative!) One argument we can add to these layers is `limits=`. This argument takes a vector (`c()` function) giving the minimum and maximum values for the axis. 

For example, below we re-create our scatterplot and change the *x*-axis to have limits from 50--90 and the *y*-axis to have limits from 0--10. We also use the `breaks=` argument to add tick marks at every two values on the *y*-axis.


```{r}
#| fig-alt: "Scatterplot displaying the relationship between life expectancy and happiness."
#| out-width: "90%"
#| message: false

#######################################
# Create scatterplot
# Change axis limits and breaks
#######################################

ggplot(data = happy, aes(x = life_exp, y = happiness)) +
  geom_point(shape = 21, color = "black", fill = "#ffde7a", size = 3) +
  labs(
    title = "LIFE EXPECTANCY AND HAPPINESS",
    subtitle = "\nCountries with higher life expectancies tend to be happier.",
    caption = "SOURCE: 2019 World Happiness Report",
    x = "Life expectancy (in years)",
    y = "Happiness (on a scale from 0-10)"
  ) +
  scale_x_continuous(
    limits = c(50, 90)
  ) +
  scale_y_continuous(
    limits = c(0, 10),
    breaks = c(0, 2, 4, 6, 8, 10)
  ) 
```

<br />



